---
title: "HW3a"
author: "Angela Larkin"
date: "10/5/2020"
abstract: "Da biggest bloop"
output: 
  word_document: 
    toc: true
    #number_sections: true
    #citation_package: pandoc-citeproc

bibliography: ["bloop.bib"]
biblio-style: "apalike"
link-citations: true
---

https://bookdown.org/yihui/rmarkdown/pdf-document.html#table-of-contents-1

https://pandoc.org/MANUAL.html#citations

https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf

https://stirlingcodingclub.github.io/Manuscripts_in_Rmarkdown/Rmarkdown_notes.html


\newpage

# Introduction

# Literature Review

# Results





```{r}
Lambda2F <- function(Lambda,dim,nuH,nuE,alpha=.05) #nuH is the number of groups - 1, nuE is number of observations - number of variables in y
{
  #dim is the number of variables you're taking out of the model
	df1 <- dim*nuH
	w <- nuH + nuE - .5*(dim+nuH+1)
	t <- 1
	if((dim^2 + nuH^2 - 5) > 0)
		t <- sqrt((dim^2 * nuH^2 - 4)/(dim^2 + nuH^2 - 5))
	df2 <- w*t - .5*(dim*nuH - 2)
	Fstat <- (1-Lambda^(1/t))/(Lambda^(1/t)) * df2/df1
	critval <- qf(1-alpha,df1,df2)
	pval <- 1 - pf(Fstat,df1,df2)
	return(list(Lambda,w,t,df1,df2,Fstat,critval,pval))
}

# sowdate:
# full: 0.00064500
# reduced: 0.00527159
lamzy <- 0.00064500/0.00527159
#fval <- Lambda2F(lamzy, 2, 3, 60-2) #incorrect
fval <- Lambda2F(lamzy, 2, 3, 60 - 12 - 2) #the last argument is n - (# of groups) - (the number of variables we are conditioning on or the # of variables in y (see slides3 slide 71))
fval[[8]] #p-value

# variety:
# full: 0.06530009
# reduced: 0.06994306
lamzy <- 0.06530009/0.06994306
fval <- Lambda2F(lamzy, 2, 3, 60-2)
fval[[8]] #p-value

# interaction:
# full: 0.13794739
# reduced: 0.21546479
lamzy <- 0.13794739/0.21546479
fval <- Lambda2F(lamzy, 2, 3, 60-2)
fval[[8]] #p-value

```
According to the above analysis, x3 and x4 are important variables to sowdate and the interaction in assessing trends but not to variety. 

#e
```{r}
#variety
e <- c(sqrt(11.896), sqrt(14.404), sqrt(13.656), sqrt(7245.6))

variety <- matrix(c(0.27131451, -0.02672873, 0.05558050, 0.00170062), ncol = 1)
diag(e) %*% variety


sowdate <- matrix(c(0.20895364,	-0.05317724,	0.17493015,	-0.00429399), ncol = 1)
diag(e) %*% sowdate

interaction <- matrix(c(0.28266108,	-0.04609473,	0.02654888,	0.00025815), ncol = 1)
diag(e) %*% interaction
```

Based on the standardized discriminant functions above (a-star), for variety, x1 is the most important. For sowdate, x1 and x3 are the most important and for the interaction x1 is the most important. @R-base 

Bloop  @R-knitr

Bloop2 @R-stringr


# Summary of Conclusions

# Future Recommendations


\newpage 



::: {#refs}
:::


```{r}
knitr::write_bib(c('knitr', 'stringr'), '', width = 60)

```


\newpage

# Appendix

```{r ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}

```





