---
title: "Title"
author: "Jared Cummings"
date: "September 2021"
abstract: "Abstract"
output: 
  pdf_document: 
    toc: true
    #number_sections: true
    #citation_package: pandoc-citeproc

#bibliography: [analysis1.bib"]
#biblio-style: "apalike"
#link-citations: true
---
```{r, setup, include=FALSE}
library(tidyverse)
farms <- read_csv('farms.csv') %>% 
  mutate(ppa = sale_price/acres/10000)
```

\newpage

# Introduction

Scientific question: Even though the land is considered equally fertile, is there a difference in sale price-per-acre between properties in Fresno and Merced? If so, which area is a better deal for passive-income investing in small-cap farmland?

Statistical hypotheses: *depends on test chosen*

Methodology: *depends on test chosen*

Alpha level: $\alpha = 0.2$.

<!-- # Literature Review -->

# Data
The data used for this analysis is an exhaustive list of farmland transactions in the cities of Fresno and Merced, California in the years 2000-2020 where less than 120 acres were exchanged. Acreage and sale price (inflation adjusted) are recorded for each sale. We investigate the price-per-acre of the sales to investigate if there is a difference in land value between these two productive farm regions of California's central valley.

```{r, boxplot, echo = FALSE, fig.cap = 'Boxplot comparison of price-per-acre of Fresno and Merced farm sales.', out.width='50%', fig.align='center'}
ggplot(farms) + 
  geom_boxplot(aes(ppa, fill = city)) +
  labs(x = 'Price-per-acre ($10k)') + 
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

```{r, summary_stats, echo = FALSE}
farms %>% 
  group_by(city) %>% 
  summarise(n = n(),
            Avg = mean(ppa),
            StDev = sd(ppa),
            Median = median(ppa),
            IQR = diff(quantile(ppa, c(0.25, 0.75)))) %>% 
  rename(City = city)# %>% knitr::kable(digits = 2, caption = 'Summary statistics of small farm sale price-per-acre in the Fresno and Merced in 2000-2020.')
```

# Methodology

## Level of significance

<!-- Why Type II error matters here -->
In the context of agriculture investing, the risk associated with incorrectly concluding that there is a preferred price location when no material difference in per-acre price exists (type I error) is entirely realized in an unnecessary limiting of perspective properties. There is not a measurable financial cost to making this mistake. The consequence of incorrectly concluding that there is no difference in per-acre price (type II error) is much more substantial. If this mistake is made there could be substantial, unnecessary overpayment without a commensurate increase in returns. In clasical hypothesis testing, the level of significance, $\alpha$, represents the probability of making a type I error. To appropriately offset risk, we adjusted $\alpha$ to control the type II error probability of making a "practically significant" over payment of 20-25% of per-acre price. In making this adjustment we align practical significance with statistical significance.



```{r, parametric_power, echo=FALSE, out.width='50%', fig.align='center'}
d <- 0.5 # corresponds to overpaying 20-25% (on average)
m <- 31 # n farms in Fresno
n <- 38 # n farms in Merced
sigma <- sd(farms$ppa)
alpha <- seq(0.0, 0.5, length.out = 100)
power <- 1 - pt(qt(alpha, m+n-2), m+n-2, d*sqrt(m*n/(m+n))/sigma)

ggplot() + 
  geom_line(aes(alpha, 1-power)) +
  labs(x = latex2exp::TeX('$\\alpha$'), y = 'Pr(Type II error)')
```

```{r}
1 - pt(qt(.2, m+n-2), m+n-2, d*sqrt(m*n/(m+n))/sigma)
```


# Results

# Summary of Conclusions

# Future Recommendations


\newpage 

<!-- bibliography will go here -->

\newpage

# Appendix
